Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-05-16 content.tex
Changes @Giappo

\begin{sidewaystable}
\centering
  \begin{tabular}{|p{3.2cm}|p{9.7cm}|p{4.5cm}@{}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Sub-argument} & 
    \textbf{Description} &
    \textbf{Possible values} \\ 
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;tree_prior; &
    Macroevolutionary diversification model &
    Yule, BD, CBS, CCP, CEP \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;clock_model; &
    Clock for the DNA mutation rates &
    strict, relaxed log-normal \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;site_model; &
    Nucleotide substitution model &
    JC, HKY, TN93, GTR \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;mutation_rate; &
    Pace at which mutation occurs &
    mutation rate $\in \mathbb{R}_{>0}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;root_sequence; &
    DNA sequence at the root of the tree &
    any combination of a, c, g, t \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;model_type; &
    Criterion to select an inference model &
    Generative, Candidate \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;run_if; &
    Condition under which an inference model is used &
    Always, Best candidate \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;do_measure_evidence; &
    Sets whether or not the evidence of the model is to be computed &
    TRUE, FALSE \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;error_function; &
    Specifies how to measure the error &
    nLTT, $|\gamma|$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;burn\_in\_fraction; &
    Specifies the percentage of initial posterior trees to discard &
    \verb;burn_in_fraction; $\in [0, 1]$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \hline
  \end{tabular}
  \caption{
    Most important parameter options.
    Yule = pure birth model (\cite{yule})
    BD = birth death (\cite{nee1994reconstructed}), 
    CBS = coalescent Bayesian skyline (\cite{drummond2005bayesian}), 
    CCP = coalescent constant-population, 
    CEP = coalescent exponential-population,
    JC = Jukes and Cantor (\cite{jukes1969evolution}), 
    HKY = Hasegawa, Kishino and Yano (\cite{hasegawa1985dating}), 
    TN93 = Tamura and Nei (\cite{tamura1993estimation}), 
    GTR = Generalized time-reversible model (\cite{tavare1986some}).
  }
  \label{tab:options}
\bigskip

  \begin{tabular}{|@{}c|p{4cm}|p{12cm}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} &
    \textbf{Macro-argument} &
    \textbf{Description} \\
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{G}$ &
    Generative model &
    The full setting to produce BEAST2 input data. 
    Its core features are the tree prior $\mathit{p_{G}}$, the clock 
    model $\mathit{c_{G}}$ and the site model $\mathit{s_{G}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  
    $\mathit{A}$ &
    Alignment model &
    It includes the parts of the generative model that directly affect the 
    alignment generation, like the clock model $\mathit{c_{G}}$ and 
    the site model $\mathit{s_{G}}$. Additional arguments can be provided, 
    such as the mutation rate and the root sequence. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{X_{i}}$ &
    $i$-th candidate experiment &
    Full setting for a Bayesian inference. It is made by a 
    candidate inference model $\mathit{I_{i}}$ and its 
    inference conditions $\mathit{C_{i}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{I}$ &
    Inference model &
    Phylogenetic inference model to run BEAST2. Likewise the generative model $G$, 
    its main components are the tree prior $\mathit{p_{I}}$, the 
    clock model $\mathit{c_{I}}$ and the site model $\mathit{s_{I}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ 
    is used in the inference. 
    They are composed by the model type, run condition and 
    wether to measure the evidence. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{E}$ & Error measure parameters & 
    Errors measurement setup that can be specified providing an 
    error function to measure the difference between the original phylogeny 
    and the inferred posterior. The initial part of the posterior that is 
    reckoned as not representative can be discarded using a burn-in fraction. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \iffalse
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure & 
      $\mathit{\tau_{p}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \fi
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in the main text and in 
    Fig~\ref{fig:pipeline}. To run the pipeline $\mathit{A}$, $\mathit{X}$ 
    and $\mathit{E}$ must be specified. Examples can be found in 
    listings~\ref{lst:create_alignment_params}, 
    \ref{lst:create_gen_experiment_explicit} and 
    \ref{lst:create_error_measure_params}.
  }
  \label{tab:definitions}
\end{sidewaystable}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
  \centering
  \begin{tabular}{ | l | l | l | }
    \hline
    \textbf{Name} & \textbf{Description} & \textbf{Listing} \\
    \hline
    \verb;pir_run; & Run \verb;pirouette; & \ref{lst:pir_run} \\
    \verb;pir_plot; & Show the \verb;pirouette; results as a plot & \ref{lst:pir_plot} \\
    \verb;create_pir_params; & Create the \verb;pirouette; parameters & \ref{lst:create_pir_params} \\
    \hline
    \verb;create_alignment_params; & Create the alignment parameters & \ref{lst:create_alignment_params} \\
    \verb;create_twinning_params; & Create the twinning parameters & \ref{lst:create_twinning_params} \\
    \verb;create_experiment; & Create one experiment & \ref{lst:create_gen_experiment_explicit} \\
    \verb;create_error_measure_params; & Create the error measurement parameters & \ref{lst:create_error_measure_params} \\
    \hline
  \end{tabular}
  \caption{
    \texttt{pirouette}'s main functions, description and the number of the 
    listing in which it is used. 
  }
  \label{tab:functions}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We show the usage of \verb;pirouette; by a story that introduces
its features gradually. First we measure the Bayesian inference error
when we have a phylogeny with a known and standard tree prior,
to get an idea of the baseline error.
Second, we'll additionally measure the inference error made by a
best candidate model, to establish that the true tree prior is indeed the best.
Lastly, we will seperate the baseline error and complete error in
a tree generated by an unknown or non-standard tree prior, 
to quantify the impact the tree prior has in Bayesian inference.

@@

\verb;pirouette; quantifies the influence of a new tree prior on BEAST2's 
inference,
by measuring the difference between a given tree and 
phylogenies in a posterior. Due to stochasticity, all posterior trees
will differ from the given phylogeny $\tau_{\mathit{G}}$, 
even when the generative tree prior $\mathit{p_{G}}$ 
and alignment model $\mathit{A}$ are the same as 
those in the inference model $\mathit{I}$.
Measuring this difference allows us to know the baseline error
of the \verb;pirouette; pipeline. We therefore define as 'standard tree priors' 
all the tree priors that are available to be used within an inference model 
(see Table~\ref{tab:options}).

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny, as described in subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists of four blocks of 250 mononucleotides each, 
where the per-nucleotide mutation rate is 0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A JC site model assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all lineages are equal and constant.

@@

As the site and clock models used here are also the defaults, 
the function arguments can be safely omitted: we just show these explicitly for clarity.

In the second step we state our experiment.
We define an experiment $\mathit{X}$ as a combination of an inference model 
$\mathit{I}$
and conditions $\mathit{C}$.
In this example we pick $\mathit{I}$ to be the same inference model as the generative one,
which is the chosen Yule tree prior $\mathit{p_{G}^0}$, as well as site and clock 
models defined in $\mathit{A}$, respectively JC and strict.
We specify in $\mathit{C}$ that the experiment will always be run.

@@

We also need to specify the error measurement parameters $\mathit{E}$.
Here we use the default error measurement
parameters, which uses a burn-in fraction of 10\% and the nLTT statistic to
measure the difference between phylogenies. For clarity,
we create this setup explicitly here:

@@

We now have create the alignment parameters,
our experiments and our error measure parameters.
These objects need to be bundled in a bigger
\verb;pirouette; parameter structure, 
using \verb;create_pir_params;:

@@

Now we can use our Yule tree and \verb;pir_params; to measure 
the inference error made on phylogenies
created by a standard diversification model:

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_1/errors.png}
  \caption{
    The inference error made 
    when the generative and inference model are the same.
    The vertical dashed line indicates the median error.
  }
  \label{fig:example_1}
\end{figure}

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we use it in the last part of this example.

@@

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$, by our definition, 
is non-standard, and hence not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is which standard tree prior should be 
used in the inference.
In this example, we re-use our hand-picked inference model,
but additionally pick a best inference model from a set of inference models.
With 'best' inference model, we mean the model with the highest 
evidence (also known as marginal likelihood), based on the alignment.
We show the procedure using \verb;pirouette; to answer a second research 
question: "What is the inference error made on a novel phylogeny when
using the best inference model, in comparison to a hand-picked model?

@@

We do need to specify a different set of experiments: we
need to state that we already have an experiment for
the generative model, as well as that we want all other inference
models to compete. We call the competing models 'candidate models',
as only the best will be used in the actual inference.
As model selection is commonly performed on the full list of available 
candidate models, \verb;pirouette; has a dedicated function for this choice:
\verb;create_all_experiments; will create a full set of 40 experiments,
containing the inference models of all combinations of 
4 site models, 2 clock models and 5 tree priors. All we need to add is
to exclude the inference model in the generative experiment:

@@

We combine the generative and all candidate models into one set of experiments:

\begin{lstlisting}[
  language = R, 
  floatplacement = ht, 
  frame = single, 
  label = {lst:create_all_experiments},
  caption = {
    Create a collection of experiments, with 1
    generative model, and 39 candidate models
  }
]
experiments <- list(
  list(generative_experiment),
  candidate_experiments
)
\end{lstlisting}

We can now create the complete
\verb;pirouette; parameter set like this (which is the
same as Listing~\ref{lst:create_pir_params}, but using the defaults):

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}),
as shown in Figure~\ref{fig:example_6}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/errors.png}
  \caption{
    The inference error
    for both the generative and best candidate inference models.
  }
  \label{fig:example_6}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparing to background noise}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Up until now, we measured the inference error on a tree
of which we knew the (standard) tree prior. The goal of \verb;pirouette;
is to measure the impact of a novel tree prior.
To do so, in this example, we'll use a tree generated by a non-standard
tree prior, of which we assume is closely related to the Yule tree prior.
To measure the impact of the tree prior, we measure both the
baseline error (that serves as a control) 
and the full inference error: their difference shows
the impact of the tree prior.
To do so, we use a twin tree (see Subsection~\ref{subsec:twinning}) for
both the generative and best candidate model.
The research question this example answers is:
"What is the inference error made from a phylogeny,
for both a generative model and a best candidate model, 
compared to the background noise?"

This time we start with a tree generated from an unknown 
diversification model, that has six taxa and a crown age of ten:
\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/true_tree.png}
  \caption{The tree derived from an unknown diversification process, 
    as created by listing~\ref{lst:unknown_phylogeny}.
  }
\end{figure}

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}),
The output is shown in Figure~\ref{fig:example_5}

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error
  }
  \label{fig:example_5}
\end{figure}

@@

To give an idea of the insights \verb;pirouette; can give,
we pretend the final result was to be obtained from thousands of replicates.
We ...

>>>>

\begin{sidewaystable}
\centering
  \begin{tabular}{|p{3.4cm}|p{9.7cm}|p{4.5cm}@{}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Sub-argument} & 
    \textbf{Description} &
    \textbf{Possible values} \\ 
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;tree_prior; &
    Macroevolutionary diversification model &
    Yule, BD, CBS, CCP, CEP \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;clock_model; &
    Clock for the DNA mutation rates &
    strict, relaxed log-normal \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;site_model; &
    Nucleotide substitution model &
    JC, HKY, TN93, GTR \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;mutation_rate; &
    Pace at which mutation occurs &
    mutation rate $\in \mathbb{R}_{>0}$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;root_sequence; &
    DNA sequence at the root of the tree &
    any combination of a, c, g, t \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;model_type; &
    Criterion to select an inference model &
    Generative, Candidate \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;run_if; &
    Condition under which an inference model is used &
    Always, Best candidate \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;do_measure_evidence; &
    Sets whether or not the evidence of the model is to be computed &
    TRUE, FALSE \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;error_function; &
    Specifies how to measure the error &
    nLTT, $|\gamma|$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \verb;burn_in_fraction; &
    Specifies the percentage of initial posterior trees to discard &
    \verb;burn_in_fraction; $\in [0, 1]$\\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \hline
  \end{tabular}
  \caption{
    Most important parameter options.
    Yule = pure birth model (\cite{yule})
    BD = birth death (\cite{nee1994reconstructed}), 
    CBS = coalescent Bayesian skyline (\cite{drummond2005bayesian}), 
    CCP = coalescent constant-population, 
    CEP = coalescent exponential-population,
    JC = Jukes and Cantor (\cite{jukes1969evolution}), 
    HKY = Hasegawa, Kishino and Yano (\cite{hasegawa1985dating}), 
    TN93 = Tamura and Nei (\cite{tamura1993estimation}), 
    GTR = Generalized time-reversible model (\cite{tavare1986some}).
  }
  \label{tab:options}
\bigskip

  \begin{tabular}{|@{}c|p{4cm}|p{12.2cm}|}
    \hline
    \centering
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \textbf{Symbol} &
    \textbf{Macro-argument} &
    \textbf{Description} \\
    \hline
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{G}$ &
    Generative model &
    The full setting to produce BEAST2 input data. 
    Its core features are the tree prior $\mathit{p_{G}}$, the clock 
    model $\mathit{c_{G}}$ and the site model $\mathit{s_{G}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%  
    $\mathit{A}$ &
    Alignment model &
    It includes the parts of the generative model that directly affect the 
    alignment generation, like the clock model $\mathit{c_{G}}$ and 
    the site model $\mathit{s_{G}}$. Additional arguments can be provided, 
    such as the mutation rate and the root sequence. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{X_{i}}$ &
    $i$-th candidate experiment &
    Full setting for a Bayesian inference. It is made by a 
    candidate inference model $\mathit{I_{i}}$ and its 
    inference conditions $\mathit{C_{i}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{I}$ &
    Inference model &
    Phylogenetic inference model to run BEAST2. Likewise the generative model $G$, 
    its main components are the tree prior $\mathit{p_{I}}$, the clock model $\mathit{c_{I}}$ and the site model $\mathit{s_{I}}$. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{C}$ & Inference conditions & Conditions under which $\mathit{I}$ 
    is used in the inference. 
    They are composed by the model type, run condition and 
    whether to measure the evidence. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    $\mathit{E}$ & Error measure parameters & 
    Errors measurement setup that can be specified providing an 
    error function to measure the difference between the original phylogeny 
    and the inferred posterior. The initial part of the posterior that is 
    reckoned as not representative can be discarded using a burn-in fraction. \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \iffalse
    $\mathit{T}$ & Twinning parameters & Twin tree creation procedure & 
      $\mathit{\tau_{p}}$ \\
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \fi
    \hline 
  \end{tabular}
  \caption{
    Definitions of terms and relative symbols used in the main text and in 
    Fig~\ref{fig:pipeline}. To run the pipeline $\mathit{A}$, $\mathit{X}$ 
    and $\mathit{E}$ must be specified. Examples can be found in 
    listings~\ref{lst:create_alignment_params}, 
    \ref{lst:create_gen_experiment_explicit} and 
    \ref{lst:create_error_measure_params}.
  }
  \label{tab:definitions}
\end{sidewaystable}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
  \centering
  \begin{tabular}{ | l | l | l | }
    \hline
    \textbf{Name} & \textbf{Description} & \textbf{Listing} \\
    \hline
    \verb;pir_run; & Run \verb;pirouette; & \ref{lst:pir_run} \\
    \verb;pir_plot; & Show the \verb;pirouette; results as a plot & \ref{lst:pir_plot} \\
    \verb;create_pir_params; & Create the \verb;pirouette; parameters & \ref{lst:create_pir_params} \\
    \hline
    \verb;create_alignment_params; & Create the alignment parameters & \ref{lst:create_alignment_params} \\
    \verb;create_twinning_params; & Create the twinning parameters & \ref{lst:create_twinning_params} \\
    \verb;create_experiment; & Create one experiment & \ref{lst:create_gen_experiment_explicit} \\
    \verb;create_error_measure_params; & Create the error measurement parameters & \ref{lst:create_error_measure_params} \\
    \hline
  \end{tabular}
  \caption{
    \texttt{pirouette}'s main functions, description and the number of the 
    listing in which it is used. 
  }
  \label{tab:functions}
\end{table}
\giovanni{I think the reference to the listings is a great idea. I don't know what to add more to this table. Probably (not 100\% sure) we should make all tables fit within the page limits.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We show the usage of \verb;pirouette; by gradually introducing
its features.
First, to get an idea of the baseline error, we measure the Bayesian inference error when we start from a phylogeny generated under a known and standard tree prior.
Second, to establish that the true tree prior is indeed the best, we also measure the inference error made adopting a
best candidate inference model.
Lastly, we quantify the impact of the tree prior in the Bayesian inference. We do so by separating the baseline error and the complete error running the pipeline from a tree generated by an unknown or non-standard tree prior.

@@

\verb;pirouette; quantifies the influence of a new tree prior on BEAST2's inference by measuring the discrepancy between a given tree and a posterior distribution of phylogenies obtained as result of the inference process. Due to stochasticity, posterior trees will generally differ from the given phylogeny $\tau_{\mathit{G}}$, even when the generative tree prior $\mathit{p_{G}}$ and alignment model $\mathit{A}$ are the same as those in the inference model $\mathit{I}$.
Measuring this difference allows us to know the baseline error
of the \verb;pirouette; pipeline. We therefore define as 'standard tree priors' all the tree priors that are available to be used within an inference model (see Table~\ref{tab:options}).

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny, as described in subsection~\ref{subsec:pipeline}.
In this example, the root sequence consists of four blocks of 250 mononucleotides each, while the per-nucleotide mutation rate is 0.1 mutations per unit time.
We use a Jukes-Cantor (JC, \cite{jukes1969evolution}) site model
and a strict clock model as these are the simplest.
A JC site model assumes that mutation rates between nucleotides are equal and constant. 
A strict clock model assumes that the mutation rates 
of all lineages are equal and constant.

@@

As the site and clock models used here are also the defaults, 
the function arguments can be safely omitted: we just explicitly show them for the sake of clarity.

In the second step we state our experiment.
We define an experiment $\mathit{X}$ as a combination of an inference model 
$\mathit{I}$
and conditions $\mathit{C}$.
In this example we pick $\mathit{I}$ to be the same inference model as the generative one,
which is the chosen Yule tree prior $\mathit{p_{G}^0}$ as well as site and clock models defined in $\mathit{A}$, respectively JC and strict.
We specify in $\mathit{C}$ that the experiment will always be run.

@@

We also need to specify the error measurement parameters $\mathit{E}$.
Here we use the default $\mathit{E}$, which uses a burn-in fraction of 10\% and the nLTT statistic to
measure the difference between phylogenies. For clarity,
we create this setup explicitly here:

@@

We now have all the needed \verb;pirouette; parameters: the alignment parameters, the experiments and the error measure parameters.
These objects need to be bundled in a bigger parameter structure, using \verb;create_pir_params;:

@@

We can finally use the given Yule tree and \verb;pir_params; to measure the inference error made on phylogenies
created by a standard diversification model:

@@

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_1/errors.png}
  \caption{
    The inference error made 
    when the generative and inference models are the same.
    The vertical dashed line indicates the median error.
  }
  \label{fig:example_1}
\end{figure}

The resulting error distribution, as shown Figure~\ref{fig:example_1},
shows the inference error 
when $\mathit{I}$ matches with the generative model given 
by $\mathit{A}$ and $\mathit{p_{G}^0}$.
This error distribution can serve as a control,
as it is obtained from a tree of which the tree prior is standard and known.
Obtaining this control is in fact an inherent part of \verb;pirouette;. 
We call it 'twinning' and we will use it starting from \ref{Comparing to background noise}.

@@

In the previous example we selected the inference model $\mathit{I}$ to match 
with a known generative tree prior $\mathit{p_{G}^0}$.
However, a novel tree prior $\mathit{p_{G}}$, by our definition, is non-standard, and hence not part of a standard inference 
model (i.e. an inference model using a standard tree prior).
In such a case, the question is which standard tree prior should be used in the inference.
In this example, not only we re-use our hand-picked inference model, but we also pick the best inference model from a set of inference models.
With the 'best' inference model, we mean the model with the highest evidence (also known as marginal likelihood), based on the alignment.
We show the procedure using \verb;pirouette; to answer a second research question: "What is the inference error made on a novel phylogeny when using the best inference model, in comparison to a hand-picked model?

@@

Here we specify a different set of experiments: we
need to state that we already have an experiment for
the generative model, as well as that we want all the other inference models to compete. We call the competing models 'candidate models'.
As model selection is commonly performed on the full list of available candidate models, \verb;pirouette; has a dedicated function for this choice: \verb;create_all_experiments; creates a full set of 40 experiments, containing the inference models of all combinations of 4 site models, 2 clock models and 5 tree priors. All we need to add is to exclude the inference model in the generative experiment:

@@

We combine the generative and all the candidate models into one set of experiments:

\begin{lstlisting}[
  language = R, 
  floatplacement = ht, 
  frame = single, 
  label = {lst:create_all_experiments},
  caption = {
    Create a collection of experiments, with 1
    generative model, and 39 candidate models.
  }
]
experiments <- list(
  list(generative_experiment),
  candidate_experiments
)
\end{lstlisting}

We can now create the complete \verb;pirouette; parameter set in the usual way (which is the same as Listing~\ref{lst:create_pir_params}, but using the defaults):

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}) in Figure~\ref{fig:example_6}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/errors.png}
  \caption{
    The inference error for both generative and best candidate inference models.
  }
  \label{fig:example_6}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparing to background noise}\label{Comparing to background noise}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

So far we have measured the inference error on a tree
generated according to a known (and standard) tree prior. The goal of \verb;pirouette; is to measure the impact of a novel tree prior.
To do so, in this example, we will use a tree generated by a non-standard tree prior, assumed to be closely related to the Yule tree prior.
To measure the impact of the tree prior, we measure both the baseline error (that serves as a control) and the full inference error: their difference shows the impact of the tree prior.
To do so, we use a twin tree (see Subsection~\ref{subsec:twinning}) for each of the generative and best candidate model.
The research question this example answers is:
"What is the inference error made from a phylogeny, for both a generative model and a best candidate model, compared to the background noise?"

We start from the tree generated by a non-standard (and unknown) tree prior, having six taxa and a crown age of ten:
\begin{lstlisting}[
  language = R, 
  floatplacement = ht,
  frame = single, 
  label = {lst:unknown_phylogeny},
  caption = A phylogeny generated by an unknown diversification model.
]
phylogeny  <- ape::read.tree(
  text = "(((A:8, B:8):1, C:9):1, ((D:8, E:8):1, F:9):1);"
)
\end{lstlisting}
\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_6/true_tree.png}
  \caption{The tree derived from an unknown diversification process, 
    as created by listing~\ref{lst:unknown_phylogeny}.
  }
\end{figure}

@@

We run \verb;pirouette; (Listing~\ref{lst:pir_run}) 
and plot the results (Listing~\ref{lst:pir_plot}).
The output is shown in Figure~\ref{fig:example_5}.

\begin{figure}[H]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    The inference error made 
    for both a generative tree prior and best candidate model
    compared with the error obtained for the twin tree.
    Here, the 'twin' tree shows the baseline inference error.
  }
  \label{fig:example_5}
\end{figure}
\giovanni{I find the figure to be too confused. What if we use instead different plots, and in each of them we just show one inference model's results plus its twin counterpart?}

@@

\iffalse
\giovanni{We can probably just remove this.}
To give an idea of the insights \verb;pirouette; can give,
we pretend the final result was to be obtained from thousands of replicates.
We ...
\fi
